NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2017-11-29 Manuscript/manuscriptPS.tex
Add separate manuscript for PS, and for PG and PE.

\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE, PG and PS lipids} %Title of paper

@@

@@

Absolute values of experimental order parameters for different lipid headgroups are
collected in Fig. \ref{HGorderParameters}. Signs are measured only for PC as far as I know,
thus only absolute values are used for now.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
  \caption{\label{HGorderParameters}
    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
  }
\end{figure}

@@

\subsection{PE headgroup}
Order parameters from Slipids simulations 
and experiments for DPPE are shown in Fig. \ref{HGorderParametersPE}.
Glycerol backbone order parameters in Slipids are off from experiments,
as already observed previously for PC lipids \cite{botan15}.
Order parameter signs for PE are not experimentally measured yet.
For headgroup the signs are set to give best agreement with simulations
and for glycerol to be consistent with experimental signs for PC.
Order parameter for $\beta$ carbon shows apparent agreement
with experiments. However, the sign of beta order parameter is positive,
in contrast to PC where negative sign was measured. Thus, the the beta order
parameter agrees with experiment with the assumption that its sign is opposite
than for PC. This is yet to be confirmed by experiments.
Order parameter for $\alpha$ carbon is too close to zero, even if the sign
would be correct.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.eps}
  \caption{\label{HGorderParametersPE}
    Order parameters for DPPE headgroup and glycerol
    backbone from simulations with Slipids \cite{??} and experiments
    (DPPE from \cite{seelig76} and E.coliPE from \cite{gally81}).
    Absolute values are shown, because signs are not known experimentally.
  }
  \todo{Experimental signs of the order parameters would be highly useful.} \\
\end{figure}

@@

\subsection{PG headgroup}
Comparison between experiments and simulations for PG lipids is shown
in Fig. \ref{HGorderParametersPOPG}. The signs are not yet measured
experimentally. They are set to give the best argeement with experiments.
This would suggest that the $\beta$ order parameter would be positive,
in contrast to PC and PS headgroups, were negative signs were measured.
Even thought the signs turned out to be correct, the tested models would
not give a very good argeement with the experiments.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
  \caption{\label{HGorderParametersPOPG}
    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
    E.Coli PG results from \cite{gally81}).
    Signs are not known for experimental data. They are determined to give best agreement
    with simulations. This is not reliable and should be corrected when experimental
    data becomes available.
  }
  \todo{More simulation data for lipids with different headgroups to be collected} \\
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.} \\
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

@@

\subsection{Effect of PE, PS and PG on PC headgroup}
The headgroup order parameters for PC lipids (POPC and DOPC)
mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
from different simulation model and experiments \cite{scherer87} with different
mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
smaller than in experiments.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPSPG.eps}
  \caption{\label{HGorderparametersPCvsPEPSPG}
    PC headgroup order parameters from mixtures with PE, PS and PG
    lipids with various mole fractions from different simulation models and experiments \cite{scherer87}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}

\subsection{Effect of PC on PS and PG headgroups}
The headgroup order parameters for PS and PG lipid mixtures with PC
having different mole fractions from simulations and experiments \cite{borle85,roux90}
are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
amount of PC to PS headgroup seems to qualitatively incorrect in CHARMM36 simulations.
The $\beta$-carbon order parameter increases in experiment, but decreases
in simulations with both tested counterions (Na+ and K+). Larger
$\alpha$-carbon order parameter decreases with the addition of PC
in experiment, while the lower remains unchanged. In simulations the
larger increases and the lower decreases.
Interestingly, the $\alpha$-carbon order parameters are closer to experiments
in pure PS system with K+ counterions than with Na+. 
The changes in PG headgroup order parameters are minor in simulations, which
is in line with the only available experiment for the $\beta$-carbon.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPSPGvsPC.eps}
  \caption{\label{HGorderparametersPSPGvsPC}
    PS and PG lipid headgroup order parameters from mixtures with PC
    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
    Signs are not yet known experimentally for PG, thus the signs give by simulations are used.
    Signs for PS are measures as described in SI.
  }
  \todo{Some simulations contain potassium as counterions, while some sodium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.} \\
  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
\end{figure*}

@@

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is linearly proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Figs \ref{OrderParametersWithCaCl},
\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \caption{\label{OrderParametersWithCaCl}
    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
    and POPC:POPG mixture data without NaCl from \cite{borle85}.
  }
  \todo{Check the NaCl concentrations in the samples.}
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaClBELOW1M}
    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters
    in the presence of different amount of negatively charged lipids
    respect to the values without added CaCl$_2$.
    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
  }
\end{figure}

>>>>

\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids} %Title of paper

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
  \caption{\label{HGorderParameters}
    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

Absolute values of experimental order parameters for different lipid headgroups are
collected in Fig. \ref{HGorderParameters}. Signs are measured only for PC as far as I know,
thus only absolute values are used for now.

@@

@@

@@

@@

\subsection{Effect of PE, PS and PG on PC headgroup}
The headgroup order parameters for PC lipids (POPC and DOPC)
mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
from different simulation model and experiments \cite{scherer87} with different
mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
smaller than in experiments.

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPEPSPG}
    PC headgroup order parameters from mixtures with PE, PS and PG
    lipids with various mole fractions from different simulation models and experiments \cite{scherer87}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure}

\subsection{Effect of PC on PS and PG headgroups}
The headgroup order parameters for PS and PG lipid mixtures with PC
having different mole fractions from simulations and experiments \cite{borle85,roux90}
are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
amount of PC to PS headgroup seems to qualitatively incorrect in CHARMM36 simulations.
The $\beta$-carbon order parameter increases in experiment, but decreases
in simulations with both tested counterions (Na+ and K+). Larger
$\alpha$-carbon order parameter decreases with the addition of PC
in experiment, while the lower remains unchanged. In simulations the
larger increases and the lower decreases.
Interestingly, the $\alpha$-carbon order parameters are closer to experiments
in pure PS system with K+ counterions than with Na+. 
The changes in PG headgroup order parameters are minor in simulations, which
is in line with the only available experiment for the $\beta$-carbon.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSPGvsPC}
    PS and PG lipid headgroup order parameters from mixtures with PC
    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
    Signs are not yet known experimentally for PG, thus the signs give by simulations are used.
    Signs for PS are measures as described in SI.
  }
  \todo{Some simulations contain potassium as counterions, while some sodium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.} \\
  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
\end{figure}

@@

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is linearly proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Figs \ref{OrderParametersWithCaCl},
\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
%  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
%    The change of PC headgroup order parameters
%    in the presence of different amount of negatively charged lipids
%    respect to the values without added CaCl$_2$.
%    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
%  }
%\end{figure}
